---
title: "HHernandez_DATA612_Project2"
author: "humbertohp"
date: "June 17, 2019"
output:
  html_document:
    highlight: pygments
    theme: cerulean
---

```{r }

```

#### Recommender System for Movies - Content-Based and Collaborative Filtering 

GroupLens Research has collected and made available rating data sets from the MovieLens web site (http://movielens.org). The data sets were collected over various periods of time. The selected dataset has ~100K movie ratings (1-5) from ~600 users on ~9000 movies. 

Users were selected at random for inclusion. All selected users had rated at least 20 movies. No demographic information is included. Each user is represented by an id, and no other information is provided.

The data are contained in the files `links.csv`, `movies.csv`, `ratings.csv` and `tags.csv`.

##### Data collection

```{r message=FALSE}

# Loading datasets, Package Installation

movies <- read.csv('https://raw.githubusercontent.com/humbertohpgit/MSDS3rdSem_DATA612/master/movies.csv')
ratings <- read.csv('https://raw.githubusercontent.com/humbertohpgit/MSDS3rdSem_DATA612/master/ratings.csv')


#install.packages("tidyverse")
library(dplyr)
library(tidyr)
```

##### Content-based Filtering

```{r message=FALSE, warning=FALSE}

# Movie Feature Matrix

# Build a matrix with genres in the columns and 0's or 1's indicating the presence of a genre in each movie

# Genres split

genres <- as.data.frame(movies$genres, stringsAsFactors=FALSE)
library(data.table)
genres_tmp <- as.data.frame(tstrsplit(genres[,1], '[|]', type.convert=TRUE), stringsAsFactors=FALSE)
colnames(genres_tmp) <- c(1:10)

genre_list <- c("Action", "Adventure", "Animation", "Children", "Comedy", "Crime","Documentary", "Drama", "Fantasy","Film-Noir", "Horror", "Musical", "Mystery","Romance","Sci-Fi", "Thriller", "War", "Western") # genres list from the dataset readme
 
genre_mat <- matrix(0,nrow(genres_tmp),length(genre_list)) #empty matrix
genre_mat[1,] <- genre_list #set first row to genre list
colnames(genre_mat) <- genre_list #set column names to genre list
 
#iterate through matrix to assign 0's or 1's accordingly
for (i in 1:nrow(genres_tmp)) {
 for (c in 1:ncol(genres_tmp)) {
  genmat_col = which(genre_mat[1,] == genres_tmp[i,c])
  if (i != nrow(genres_tmp)){
  genre_mat[i+1,genmat_col] <- 1
  }
 }
}
 
genre_mat <- as.data.frame(genre_mat[-1,], stringsAsFactors=FALSE) #remove first row, which was the genre list
for (c in 1:ncol(genre_mat)) {
  genre_mat[,c] <- as.numeric(genre_mat[,c])
}

head(genre_mat,5)

# User Profile Matrix

# Build a matrix with userd in the columns, movies as rows and the ratings set to 0 when there are NAs

#Simplify ratings values to a binary scale

bin_ratings <- ratings
for (i in 1:nrow(bin_ratings)){
 if (bin_ratings[i,3] > 3){
   bin_ratings[i,3] <- 1
 }
 else{
   bin_ratings[i,3] <- -1
 }
}

user_mat <- spread(bin_ratings[,1:3], userId, rating)

for (i in 1:ncol(user_mat)){
  user_mat[which(is.na(user_mat[,i]) == TRUE),i] <- 0
}
user_mat = user_mat[,-1]

head(user_mat,1)

# Matrices Dot Product - User preference towards movie genres

dim(genre_mat)
dim(user_mat)

#Remove movies that have no rating in the dataset

movieIds <- unique(movies$movieId) #9741
ratingmovieIds <- unique(ratings$movieId) #9724

genre_mat_final <- genre_mat[-which((movieIds %in% ratingmovieIds) == FALSE),]
rownames(genre_mat_final) <- NULL

genre_mat_final <- as.data.frame(genre_mat_final, stringsAsFactors=FALSE)
for (c in 1:ncol(genre_mat_final)) {
  genre_mat_final[,c] <- as.integer(genre_mat_final[,c])
}

user_mat_final <- user_mat[-9724,]

dim(genre_mat_final)
dim(user_mat_final)

dot_prod <- matrix(0,18,610)
for (c in 1:ncol(user_mat_final)){
  for (i in 1:ncol(genre_mat_final)){
    dot_prod[i,c] <- sum((genre_mat_final[,i]) * (user_mat_final[,c]))
    ## genre_mat_final[,i] %*% user_mat_final[,c]
  }
}

dot_prod[,1]

#Simplify to a Binary scale

for (c in 1:ncol(dot_prod)){
  for(i in 1:nrow(dot_prod)){
    if (dot_prod [i,c] > 0){
      dot_prod [i,c] <- 1
    }
    else {
      dot_prod [i,c] <- 0
    }
  }
}

# Similarity Calculation between user profiles and movie genres

# Calculation for 3 users (optimizing processing time)

recomm_movies_5users <- list()

for (i in 1:5){
  result_user <- dot_prod[,i] # Specific user's profile
  sim_mat <- rbind.data.frame(result_user, genre_mat_final)
  sim_mat <- data.frame(lapply(sim_mat,function(x){as.integer(x)}))
 
#Calculate Jaccard distance between user profile and all movies

  library(proxy)
  sim_results <- dist(sim_mat, method = "Jaccard")
  sim_results <- as.data.frame(as.matrix(sim_results[1:9723]))
  rows <- which(sim_results == min(sim_results))

  #Top Recommended movies for specific (5) users
  recomm_movies_5users[[i]] <- list(movies[rows,2])
}

# Top Recommended for Users (1-5)
recomm_movies_5users
```

##### User-Based Collaborative Filtering Approach

```{r message=FALSE, warning=FALSE}

# Build a user matrix with movies as columns

rating_mat <- spread(ratings[,1:3], movieId, rating)
rating_mat <- as.matrix(rating_mat[,-1]) #remove userIds

library(recommenderlab)

#Convert into a recommenderlab sparse matrix
rating_mat <- as(rating_mat, "realRatingMatrix") 
 
#Normalize data
rating_mat_norm <- normalize(rating_mat)
 
#Create Recommender Model. "UBCF" (User-Based Collaborative Filtering)

recommender_model <- Recommender(rating_mat_norm, method = "UBCF", param=list(method="Cosine",nn=30))

#Obtain top 10 recommendations for users (1-5)
recom <- predict(recommender_model, rating_mat[1:5], n=10) 
recom_list <- as(recom, "list") #convert recommenderlab object to readable list
 
recom_result <- list()

for (i in c(1:5)){
 recom_result[[i]] <- movies[as.integer(recom_list[[i]]),2]
}

# Top 10 recommendations for users (1-5)
recom_result

```

##### Conclusion: Collaborative Filtering vs Content Based

Collaborative Filtering: correlating personal preferences

- If User1 loves A and B, and User2 loves A, B, and C, then User1 is more likely to love C 
- Discover patterns in observed preference behavior across community of users
- Predict new preferences based on those patterns 
- Does not rely on item or user attributes/features

Content Based Filtering: understanding user / item profiles

- User1 is in zip code A and age group N and bought item X; User2 in same area and age group is likely to like X
- Trained on user features and item features to predict preferences based on those patterns
- Can handle cold starts (new users / items with no interaction data)

