---
title: "HHernandez_DATA612_Project4"
author: "humbertohp"
date: "June 28, 2019"
output:
  html_document:
    highlight: pygments
    theme: cerulean
---

```{r }

```

#### Recommender System Evaluation - Accuracy and Beyond

Recommendation system for music artists based on user preferences according to the songs listening frequencies (called weights) used as ratings

This dataset contains social networking, tagging, and music artist listening information from a set of 2K users from Last.fm online music system.
The dataset is released in the framework of the 2nd International Workshop on Information Heterogeneity and Fusion in Recommender Systems (HetRec 2011)  http://ir.ii.uam.es/hetrec2011 
http://www.last.fm 

Data Statistics:
1892 users
17632 artists
92834 user-listened artist relations, i.e. tuples [user, artist, listeningCount]

##### Data collection, exploration and preparation

```{r message=FALSE, warning=FALSE}

# Loading datasets, Package Installation

artists <- read.delim('https://raw.githubusercontent.com/humbertohpgit/MSDS3rdSem_DATA612/master/artists.dat', header = TRUE, sep="\t")
artists <- artists[,1:2]
user_artists <- read.delim('https://raw.githubusercontent.com/humbertohpgit/MSDS3rdSem_DATA612/master/user_artists.dat', header = TRUE, sep="\t")

#install.packages("tidyverse")
#install.packages("recommenderlab")
#install.packages("recosystem")
library(dplyr)
library(tidyr)
library(recommenderlab)
library(recosystem)

#Exploring data sets Transforming weights to ratings
head(artists)
head(user_artists)

# Reduce data set to the most relevant users (with minimum listening frequency = 10) and the top artists based on total listening frequency

user_artists_mat <- user_artists %>% spread(key=artistID,value=weight)

artist_freq <- colSums(user_artists_mat[,-1],na.rm = TRUE)

top_user_artist <- user_artists_mat[,order(artist_freq,decreasing = TRUE)[1:100]]
list_freq <- apply(top_user_artist,1,function(x) return(sum(!is.na(x))))
top_user_artist <-  top_user_artist[list_freq > 10,]
dim(top_user_artist)

# Transforming weights to ratings

quantile(user_artists$weight, na.rm=TRUE) #determine ratings limits
quantile(user_artists$weight, probs=0.90,na.rm=TRUE)

user_artist_trans <- top_user_artist
user_artist_trans[user_artist_trans$weight < 107] <- 1
user_artist_trans[user_artist_trans >= 107 & user_artist_trans < 260] <- 2
user_artist_trans[user_artist_trans >= 260 & user_artist_trans < 614] <- 3
user_artist_trans[user_artist_trans >= 614 & user_artist_trans < 1387] <- 4
user_artist_trans[user_artist_trans >= 1387] <- 5

head(user_artist_trans,3)

# Create sparse matrix and the corresponding Train and Test data sets

user_artist_trans_mat <- as(as.matrix(user_artist_trans),"realRatingMatrix")
user_artist_trans_mat

set.seed(100)
eval_sch <- evaluationScheme(data = user_artist_trans_mat, method = "split", train = .8, given = 3, goodRating = 3)

```

##### recommenderlab - Different Recommender Methods

```{r message=FALSE, warning=FALSE}

#Exploring parameters of recommendation methods in recommenderlab

recommender_models <- recommenderRegistry$get_entries(dataType = "realRatingMatrix")
names(recommender_models)
lapply(recommender_models, "[[", "description")

# Create recommender model - User-based Collaborative Filtering (UCBF)

recommender_model_ubcf <- Recommender(data=getData(eval_sch,"train"),method="UBCF")

# Recommendation for the first 5 users

recom_ubcf <- predict(recommender_model_ubcf, getData(eval_sch,"known")[1:5], n=5, type = "topNList")
recom_ubcf_list <- as(recom_ubcf, "list") #convert recommenderlab object to readable list
recom_ubcf_list 

# Get artist names

recom_ubcf_result <- list()
for (i in c(1:5)){
 recom_ubcf_result[[i]] <- artists[artists$id %in% as.vector(recom_ubcf_list[[i]]),2]
}

# Top 5 recommendations for the first 5 users
recom_ubcf_result

# Evaluation

# Prediction of ratings for all users

recom_ubcf <- predict(recommender_model_ubcf, getData(eval_sch,"known"), type = "ratings")

accuracy_ubcf <- calcPredictionAccuracy(x = recom_ubcf, data = getData(eval_sch, "unknown"), given=3, goodRating=3)

accuracy_ubcf

# Create recommender model - Alternating least squares (ALS)

recommender_model_als <- Recommender(data=getData(eval_sch,"train"),method="ALS")

# Recommendation for the first 5 users

recom_als <- predict(recommender_model_als, getData(eval_sch,"known")[1:5], n=5, type = "topNList")
recom_als_list <- as(recom_als, "list") #convert recommenderlab object to readable list
recom_als_list 

# Get artist names

recom_als_result <- list()
for (i in c(1:5)){
 recom_als_result[[i]] <- artists[artists$id %in% as.vector(recom_als_list[[i]]),2]
}

# Top 5 recommendations for the first 5 users
recom_als_result

# Evaluation

# Prediction of ratings for all users

recom_als <- predict(recommender_model_als, getData(eval_sch,"known"), type = "ratings")

accuracy_als <- calcPredictionAccuracy(x = recom_als, data = getData(eval_sch, "unknown"), given=3, goodRating=3)

accuracy_als

```

##### recommenderlab - Recommender Methods Comparison

```{r message=FALSE, warning=FALSE}

comp_methods <- rbind(accuracy_ubcf, accuracy_als)
rownames(comp_methods) <- c("UBCF", "ALS")
library(knitr)
kable(comp_methods)

# ROC & Precision & Recall

rec_methods <- list(
  "UBCF" = list(name = "UBCF", param = NULL),
  "ALS" = list(name = "ALS", param = NULL))
eval_results <- evaluate(x = eval_sch, method = rec_methods, n = c(1, 5, 10, 15, 20,25,30))

plot(eval_results, 
     annotate = TRUE, legend = "topleft", main = "ROC Curve")

plot(eval_results, "prec/rec", 
     annotate = TRUE, legend = "topright", main = "Precision-Recall")

```

##### Observations

In terms of model method accuracy, Alternate Least Square (ALS) provides better RMSE and MSE than the UBCF.
Interestingly in the ROC curve, UBCF appears to perform better that ALS, meaning that in general the classification power of UBCF is better even though that in accuracy terms, ALS, shows less deviation from the actual rating values when it classifies correctly



